---
title: "BIOS611Project"
author: "Kevin Wang"
date: "2022-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#library(reticulate)
library(tidyr)
#library(matlab)
library(MASS)
library(stats)
# library(nlme)
library(tibble)
library(dplyr)
library(data.table)
# library(assist)
library(mgcv)
library(ggplot2)
# library(lme4)
#library(kernlab)
#library(plotly)
# library(numDeriv)
# library(randtoolbox)
# library(mrgsolve)
# library(parallel)
# library(BSDA)
# library(epitools)
#library(rdist)
#library(gbm)
```

# Getting data


```{r secure the (data)}
data = read.csv("brain_stroke.csv")
summary(data)
```

# Exploratory Data analysis


```{r exploratory and cleaning}
#plot(data)
data$smoking_status = as.factor(data$smoking_status)
data$work_type = as.factor(data$work_type)
data$Residence_type = as.factor(data$Residence_type)
data$gender = as.factor(data$gender)
data$ever_married = as.factor(data$ever_married)
summary(data)

numerics = c("age","hypertension", "heart_disease", "avg_glucose_level", "bmi","stroke")
cor(data[,numerics]) %>% round(digits = 2)
```


```{r naieve model}
fit1 = glm(stroke ~ .-stroke, family = binomial, data = data)
summary(fit1)

fitbase = glm(stroke ~ age+hypertension+avg_glucose_level, family = binomial, data = data)
summary(fitbase)

anova(fitbase,fit1, test = "Chisq")
#anova(fitbase,fit1, test = "F")
```

Now that we have a little idea about what matters, we can try some slightly more complicated models before we move on to the clustering part of our EDA.

Before, the significant variables were age, hypertension, and average glucose (and the intercept). We can try to model some components non-linearly and see if any others become significant. This doesn't give us any conclusions about the data, just gives us ideas on where to look next in our exploratory data analysis. The actual models for drawing conclusions will come later.


```{r}
paste(names(data), collapse=" + ")
fit2 = gam(stroke ~ 
             gender + s(age) + hypertension + heart_disease + ever_married +
             work_type + Residence_type + s(avg_glucose_level) + s(bmi) + smoking_status
, family = binomial, data = data)

summary(fit2)
```

Deviance explained is quite low, so either our covariates are not very informative or we are missing things in our model. Its most likely the latter, meaning theres a lot more EDA to do. We will also investigate the possibility of some interactions using the tensor product splines. This is again not a rigorous analysis, just some hints towards where the data might lead us later.


```{r}
fit_interactions = gam(stroke ~ 
             s(age) + s(avg_glucose_level) + s(bmi) + ti(age,avg_glucose_level) + 
             ti(age,bmi)+ti(avg_glucose_level,bmi)
, data = data)


summary(fit_interactions)
```
There do appear to be interactions, which we might be able to further investigate when we cluster. Hopefully, we'll be able to see whats actually going on when we plot the data.

Summary: age, glucose, and hypertension appear to be significant. Work type may be significant, but only depending on how we treat the other variables, indicating there may be a complicated interaction. Will need to visualize. BMI does not appear to have a significant effect, but this may be due more to covariance. The remaining variables were not found to add significant information to our naive models. We will still investigate them, but we will not make too much trouble if they aren't interesting.


```{r}
names(data)
```

```{r}
tiff("outputs/scatters.tiff")
plot(data)
dev.off()

tiff("outputs/byage.tiff")
boxplot(data$age~data$stroke)
dev.off()

tiff("outputs/byglucose.tiff")
boxplot(data$avg_glucose_level~data$stroke)
dev.off()

tiff("outputs/bybmi.tiff")
boxplot(data$bmi~data$stroke)
dev.off()


numdata = data[,numerics]
pca = prcomp(numdata)
plot(pca$sdev)

tiff("outputs/variancebycomponents.tiff")
plot(pca$sdev)
dev.off()


library("Rtsne")
tiff("outputs/tsne.tiff")
out = Rtsne(numdata,pca=TRUE)
plot(out$Y)
dev.off()

```

